---
title: "TimeResolvedMR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TimeResolvedMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# TimeResolvedMR

The `TimeResolvedMR` package allows you to estimate time-varying effect of an
exposure on an outcome using genetic instrumental variables (IVs) through
Mendelian Randomization (MR). At the moment, this package is tested for
continuous exposures and binary outcomes.

## Data Preparation

You require the following data for your analysis:

-   Polygenic scores of the exposure for each individual as IV. This is assumed
    to be correct and no checks will be performed. (`pgs`, numeric vector)
-   Exposure measurements (`exposure`) and the age at which the exposure was
    assessed for each individual (`exposure_age`) (both numeric vectors).
-   Outcome status (`outcome`) and age at which the outcome occurred
    (`outcome_age`) (both numeric vectors)
-   Covariates (`covariates`, data.frame)

```{r data}
# Random example data for 1000 individuals
pgs <- rnorm(1000)
exposure <- pgs * 3
exposure_age <- rep(40:70, length.out = 1000)
covariates <- data.frame(covar = rnorm(1000))
outcome <- sample(c(TRUE,FALSE), 1000, replace=TRUE)
outcome_age <- rnorm(1000, mean=60, sd = 3) |> pmax(40)
```

## All-in-one Analysis

The package comes with a convenience function performing an all-in-one analysis.
This allows you to quickly run an MR analysis but you will not be able to reuse
any of the computed models.

The following code runs the MR analysis with default settings, i.e. the age-pgs
interaction is estimated using a LOESS model and the genetic effects on the
outcome using an Aalen model.

```{r all_in_one}
library(TimeResolvedMR)

all_in_one <- analyze_all_in_one(pgs, exposure, outcome, exposure_age,
                                 outcome_age, covariates)
head(all_in_one)
```

You can change the ages at which to estimated pgs effects for the LOESS model
using `age_range` and `age_step`.
Changing the step size might produce erroneous results, however.
```{r custom1, warning=FALSE}
all_in_one <- analyze_all_in_one(pgs, exposure, outcome, exposure_age,
                                 outcome_age, covariates,
                                 age_range = c(50,70), age_step = 5)
head(all_in_one)
```

By setting the `exposure_modeltype` to `"glm"` you can run a generalized linear
model that includes interaction terms between PGS and age and allows you to
model non-linear relationship through the `interaction_exponents` parameter.

```{r custom2}
all_in_one <- analyze_all_in_one(pgs, exposure, outcome, exposure_age,
                                 outcome_age, covariates,
                                 exposure_modeltype = "glm",
                                 interaction_exponents = c(1,4))
head(all_in_one)
```

These functions will be discussed in more detail below.

## Running Models Yourself

`TimeResolvedMR` also allows you to estimate time-dependent effects on exposure
and outcome separately. This way, you can reuse and mix and match models for
your analyses.

These are the functions and models that come with `TwoSampleMR` including what
they can be used for:

| Function               | Model | Response Type | Exposure | Outcome |
|------------------------|-------|---------------|----------|---------|
| `time_dependent_loess` | LOESS | Continuous    | ✅       | ❌      |
| `time_dependent_glm`   | GLM   | Continuous    | ✅       | ❌      |
| `time_dependent_aalen` | Aalen | Binary        | ⚠️       | ✅      |

✅
= Tested and intended use
⚠️
= Usable in theory but not tested
❌
= not supported

### LOESS Model
The LOESS model estimates time-dependent genetic effects by stratifying the
data by age, estimating the effect using a glm in each stratum and then
modeling the effect of the age on the genetic effect using a LOESS model.

A basic call looks like this:
```{r loess1}
loess_model <- time_dependent_loess(pgs, exposure, exposure_age, covariates)
```
The ages at which to estimate the genetic effect are automatically detected.
They are, by default, each year from the lowest to the highest value in
`exposure_age`.
You can change this behavior by specifying `age_range` and `age_step`.

The following code will estimate the effect for every other year between 50 and
70:
```{r loess2}
loess_model <- time_dependent_loess(pgs, exposure, exposure_age, covariates,
                                    age_range = c(50,70), age_step = 2)
```


### Generalized Linear Model
The glm implementation estimates time-dependent effects by running a glm
with PGS-age interaction terms.

```{r glm1}
glm_model <- time_dependent_glm(pgs, exposure, exposure_age, covariates)
```
This will include a single linear interaction term, i.e., `PGS:age`

You can include multiple interaction terms and terms with non-linear
interactions using the `exponents` parameter.

```{r glm2}
glm_model <- time_dependent_glm(pgs, exposure, exposure_age, covariates, exponents = c(1,4))
```
This model has a linear and a quartic interaction term,
i.e., `PGS:age + PGS:(age^4)`.

Setting `exponents` to `NULL` or an empty vector will create a model without
interaction terms.
This can be useful if you want to compare a time-dependent model to a
time-constant one.

### Aalen Model
The Aalen additive hazards model estimates time-dependent effects on a binary
response.

```{r aalen1}
aalen_model <- time_dependent_aalen(pgs, outcome, outcome_age, covariates)
```
At this point, there are no customization options for this model.

## Performing the Mendelian Randomization
The MR analysis estimates the effect of the exposure on the outcome depending on
the duration of the exposure.
This is done through numeric integration of the time-dependent effects at
pre-determined points.
Therefore, you must specify the points at which to estimate the effect.
```{r}
mr_results <- time_dependent_MR(age_seq = 0:80, exposure_model = glm_model,
                                outcome_model = aalen_model)
head(mr_results)
```
By default, the effects are integrated using the trapezoidal rule but you can
choose to use the midpoint rule instead.
```{r}
mr_results <- time_dependent_MR(age_seq = 0:80, exposure_model = glm_model,
                                outcome_model = aalen_model, method = "midpoint")
head(mr_results)
```

This method returns a `data.frame` containing the ages at which the effects were
estimated, the exposure's cumulative effect on the outcome (`Gamma`), its
variance and 95% confidence interval at each age.
Note that the trapezoidal rule estimates the cumulative effect at points halfway
between the ages specified in `age_seq` whereas the midpoint rule estimates them
exactly at the ages you specify.

## Adding Models
`TimeResolvedMR` is designed to be extended.
If you would like to implement another kind of time-dependent model, you can do
so by extending the defining an S4 class extending `TimeDependentModel`.
Requirements of your new class are that it includes a slot called `model`, which
contains the underlying statistical model (similar to e.g. `glm`) and implements
the method `totalEffect(this, age)`, which returns the total genetic effect at
a certain time point (_not_ the cumulative effect up until that point).
A model that can be used for genetic effects on an outcome also must implement
`variance(this, age)`, which returns the effect estimate's variance at a
certain age.

For instance, a class encapsulating a Cox proportional hazards model might look
like this:
```{r extension}
# coxph is an S3 class. Make it visible to the S4 system
setOldClass("coxph")

CoxTimeDependentModel <- setClass(
  "CoxTimeDependentModel",
  contains = "TimeDependentModel",
  slots = c(
    model = "coxph"
    # Additional slots as needed
  )
)

setMethod("totalEffect", "CoxTimeDependentModel", function(this, age){
  # Implementation goes here
})
```
This is just an example skeleton.
A functional implementation is left as an exercise for the reader.
